Modeling the Java Card API (or Interface?) in B

### 2012-03-29 trunk/papers/B2012-work/javacardAPIpaper.tex
incluido texto sobre overloading

%\lstset{language=B, caption={B machine Buscard},label=lis:addimplementation}
%\begin{lstlisting}
%	add(apdu) =
%	BEGIN
%		buffer <- - getBuffer();
%		credit <-- cast ( );
%		IF 
%			not(credit >= 0)      !!! byte version
%		THEN
%			ISOException.throwIt(EXCEPTIONS.negative_credit)
%		ELSIF 
%			not(balance + credit <= MAXIMUM)     !!! byte version
%		THEN
%			ISOException.throwIt(EXCEPTIONS.balance_exceeded)
%		ELSE
%		        balance := sum_byte (balance, credit)  
%		END
%	END
%	
%	[[[[BRUNO e SIMONE: isso esta todo errado - coloquei aqui mais ou menos o que eu acho que deve ser a cara da implementacao B que gera o codigo java card. Voces devem ter algum exemplo para recuperar a estrutura certa, nao?]]]]
%	
%\end{lstlisting}

@@

The transition into a defensive style of programming is part of the BSmart method, and implemented by the BSmart Full Function component. This component generates a B refinement of the original specification where pre-conditions are transformed into conditionals (IF-THEN-ELSE-END constructs), with the indication of the need to throw the corresponding exception when conditions are not met by input data and states.
The resulting B model can then be formally verified as usual in the B method.

@@

But all of these transformations rely, at the bottom, on the need to reference Java Card API types, classes and methods on the B level.
This is then the reason why the Java Card API needed to be the first Java Card component to be modeled in B.

@@

The Java Card API, in its version 2.2.2, is composed of 14 packages,
containing 93 classes and interfaces, as shown in
Table~\ref{tab:packages}. Their interfaces have all been modeled in B,
and the resulting models are available at the KitSmart project
page\footnote{https://code.google.com/p/kitsmart/}. Differently from
JML and OCL, though, B is not object oriented, and some modeling
solutions had to be found to cope with this lack of resources.  We
present in the following the main characteristics and restrictions of
our models:

@@

\item Conditions on method inputs which lead to exception raising in
  the Java Card platform specification are coded into pre-conditions
  in B.

  [[[ AMM (em que casos isso não é feito - NULL? outros?) ]]]

\item Method outputs are modeled in B as taking any value of the
  correct type

  [[[ AMM (e em que casos temos um pouco mais). ]]]

@@

\end{itemize}

As previously stated,  we chose to model exception raising conditions as operation pre-conditions in the B models. For instance, if the Java Card specification states that a method should throw an exception in case a parameter array is empty, the pre-condition of the corresponding B operation will require that this parameter is not empty. The B method then requires that every call to that operation satisfies the operation's pre-condition. As a consequence, code that is formally generated using the B method and the Java Card API specification should never raise this specific exception, as it is statically proved that all calls to this operation are made with non-empty parameter arrays. Not all situations leading to exceptional conditions are modeled, however, as some of them depend on the Java Card Runtime Environment (JCRE) and its state, which has not yet been modeled, and some require a more detailed specification of the contents of some data and other internal behavior.

>>>>

@@

The transition into a defensive style of programming is part of the BSmart method, and implemented by the BSmart Full Function component. This component generates a B refinement of the original specification where pre-conditions are transformed into conditionals (IF-THEN-ELSE-END constructs), with the indication of the need to throw the corresponding exception when the specified conditions are not met by input data and states.
The resulting B model can then be formally verified as usual in the B method.

@@

But all of these transformations rely on the need to reference Java Card API types, classes and methods on the B level.
This is then the reason why the Java Card API needed to be the first Java Card component to be modeled in B.

@@

The Java Card API, in its version 2.2.2\footnote{Java Card 2.2.2, although not the most recent version of the Java Card platform, is still widely used, and can easily be extended to Java Card 3 Classic Edition,  a backward compatible evolution of version 2.2.2., including some bug fixes, clarifications against the Java Card v2.2.2 specifications and new security algorithms (Java Card 3 Platform Release Notes, available at http://java.sun.com/javacard/3.0.1/specs.jsp).}, is composed of 14 packages, containing 93 classes and interfaces, as shown in
Table~\ref{tab:packages}. Their interfaces have all been modeled in B, and the resulting models are available at the KitSmart project
page\footnote{https://code.google.com/p/kitsmart/}. 
Differently from JML and OCL, though, B is not object oriented, and some modeling solutions had to be found to cope with this lack of resources.  We present in the following the main characteristics and some restrictions of our models:

@@

\item Conditions on method inputs which may lead to exception raising in the Java Card platform specification are coded into pre-conditions
  in B.

 \item Method outputs are modeled in B as taking any value of the correct type, or, when possible in the current abstraction level, any value of a more restricted set of specified return values.

@@

\item Overloading is not allowed in B and occurs frequently in Java Card. To model overloaded methods in B an extension to overloaded method names was included. We then have equals, equals\_1, equals\_2, etc. 
\end{itemize}

As previously stated,  we chose to model exception raising conditions as operation pre-conditions in the B models. For instance, if the Java Card specification states that a method should throw an exception in case a parameter array is empty, the pre-condition of the corresponding B operation will require that this parameter is not empty. The B method then requires that every call to that operation satisfies the operation's pre-condition. As a consequence, code that is formally generated using the B method and the Java Card API specification should never raise this specific exception, as it is statically proved that all calls to this operation are made with non-empty parameter arrays. Not all situations leading to exceptional conditions are modeled, however, as some of them depend on the Java Card Runtime Environment (JCRE) and its state, which has not yet been modeled, and some require a more detailed specification of the contents of some data and other internal behavior. In these cases, the  exception raising conditions are only left as comments to the B specification, and can be used as additional documentation or to be considered in future work.

One delicate issue to be dealt with both by the user of the API specification and the tools designed to work with it, is the use of overloaded methods in an application development. When overloaded operations are used in an application development, the user must explicitly choose which version of the overloaded operation is to be used. Then, verifications will be carried out considering this specific version of the operation. When code is generated, either the code generator will remove the name extension automatically, if designed for it, or the user will have to edit the generated code manually. In any case, it is important that the actual method called in the Java Card code is the one with which verifications have been carried out.   In case of overriding, in particular, it will the user's responsibility to choose in B the one which is on scope at this specific point in Java, unless a extra verification step is included in the code generator that checks that the operation used in B is the one in scope in Java and registers an error in case it is not.
